---
title: "Time Series - Homework 3"
author: "Brandon Fenton and Allison Theobold"
date: "September 21, 2016"
output:
  pdf_document:
    fig_height: 3
    fig_width: 5
  html_document:
    fig_height: 3
    fig_width: 5
  word_document:
    fig_height: 3
    fig_width: 5
header-includes: \usepackage{float} \usepackage{bm} \usepackage{amsmath} \usepackage{amssymb} \usepackage{microtype}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(warning = FALSE)

library(pander)
library(dplyr)
library(effects)

panderOptions('missing', "-")

pander_lm <-function (fit, ...)
{
  fit.sum <- summary(fit)
  fit.coef <- fit.sum$coefficients
  fit.ttable <- matrix(nrow=length(fit.sum$aliased), ncol=4)
  colnames(fit.ttable) <- colnames(fit.sum$coefficients)
  rownames(fit.ttable) <- names(fit.sum$aliased)

  notna <- as.vector(which(!fit.sum$aliased))
  fit.ttable[notna,] <- fit.coef
  fit.ttable <- as.data.frame(fit.ttable)
  fit.ttable$`Pr(>|t|)` <- ifelse(fit.ttable$`Pr(>|t|)` < 0.0001, "<0.0001",
                                     sprintf("%.4f", fit.ttable$`Pr(>|t|)`))
  

  pander(fit.ttable, ...)
}

pander_anova <-function (fit, ...)
{
  fit.anova <- anova(fit)
  fit.anova$`Pr(>F)` <- ifelse(fit.anova$`Pr(>F)` < 0.0001, "<0.0001",
                                  sprintf("%.4f", fit.anova$`Pr(>F)`))

pander(fit.anova, ...)
}
```


##1. Provide a reason for your choice of location. Report any missing observations and the range of years where you are modeling.
    
Barrow is a location of personal interest since it's in Alaska, and it's a location with fairly extreme weather so there might be interesting patterns in the CO2 concentrations relative to other locations.  There are no missing observations in the data set, but 1972 is the first year with observations for each month, so the period from January 1972 to December 2015 will be used for modeling:

```{r p1, cache=T}
BRW_flask<-tbl_df(read.table("BRW_flask.txt",header=T)) 

pander(summary(BRW_flask), caption="Summary statistics for Barrow data")

year_tab <- matrix(c(NA,table(BRW_flask$year),rep(NA,4)), nrow=5, ncol=10, byrow = TRUE)
colnames(year_tab) <- pandoc.strong.return(0:9)
rownames(year_tab) <- c(paste(197:199,"-", sep=""),paste(200:201,"-", sep=""))
pander(year_tab, caption = "Measurement counts by year", split.table=Inf)

BRW_flask <- filter(BRW_flask, year>=1972)

BRW_flask <- BRW_flask %>% mutate(month = as.factor(month)) %>%
  mutate(year = as.vector(time(ts(BRW_flask$value,start=c(1972,1),freq=12))))

```

Brandon, you get to describe why you picked this site! The table above displays the number of observations per year from 1971 to 2015. It appears 1971 is the only year with ``missing" observations, so the data being considered begins in January of 1972. 



##2. Make a nice looking time series plot of the CO$_2$ concentrations.

```{r p2, cache=T}

plot(value ~ year, data=BRW_flask, type="l", ylab=expression(CO[2] (micromol/mol)))
```

##3. Fit a linear trend plus seasonal means model to the data. Report and discuss the four panel residual diagnostics. Also make a plot of residuals vs time and discuss any potential missed pattern versus time.

```{r p3, fig.height=6, fig.width=6}

# Just year and month first to determine seasons
ym_lm <- lm(value ~ year + month, data = BRW_flask)

# plot(allEffects(ym_lm))

par(mfrow = c(2, 2))
plot(ym_lm)
par(mfrow = c(1,1))

```

 In the above diagnostic plots, we see very clear indications of assumption violations. In the residual versus fitted, we see evidence of a quadratic relationship that was unaccounted for. In the location-scale plot, the curvature appears more quartic, leading us to believe a polynomial of degree greater than 2 should possibly be fit. We see some shorter tails in the Normal Q-Q plot of the residuals, but nothing too severe to violate the assumption of normality of the residuals. Lastly, we see that there are no high leverage points in the data set. 
 
```{r, p3aaa}
plot(residuals(ym_lm) ~ year, data = BRW_flask)
``` 

In the above plot of the residuals from the linear model, with only year and month as explanatory variables versus year, shows a similar trend to that seen in the scale-location diagnostic plot. We see evidence of a potential quartic trend as discussed previously. 

```{r p3aa}

# Dec-Jan and then Jun-Nov?
BRW_flask <- BRW_flask %>% mutate(season = as.factor(ifelse(BRW_flask$month %in% c(12,1:5), 1, 2)))

ltsm_lm <- lm(value ~ year + season, data=BRW_flask)
plot(allEffects(ltsm_lm))
```

```{r p3a, fig.height=6, fig.width=6}
par(mfrow=c(2,2))
plot(ltsm_lm)
par(mfrow=c(1,1))
```

```{r p3b}
plot(resid(ltsm_lm) ~ BRW_flask$year )
```

##4. Provide tests for the linear and seasonal means components, conditional on each other. Report those results in two sentences including all details.

```{r p4}
pander_anova(ltsm_lm)

```

##5. For your model, plot the original time series and the model fitted values, both versus time on the same plot. You might consider two line types or colors for the two lines. The easiest way to obtain fitted values in R is using fitted(modelname). Discuss how it appears your model does or does not describe the responses using this plot.

```{r p5}
colors <- c("#FF0000DD", "#000000DD")
plot(value ~ year, data=BRW_flask, type="l", lty=1, lwd=2, col=colors[1], ylab=expression(CO[2] (micromol/mol)))
lines(fitted(ltsm_lm) ~ BRW_flask$year, lty=2, col=colors[2], lwd=2)
legend(x=min(BRW_flask$year), y=max(BRW_flask$value), lty=c(1,3), lwd=c(2,2), col=colors, legend=c("Original","Fitted"),
       bty = "n")
```

##6. Document your R version 

```{r}
getRversion()
```

\pagebreak

## R Code Appendix:

Setup:
```{r sup, ref.label='setup', eval=F, tidy=TRUE}

```

Problem 1:
```{r a1, ref.label='p1', eval=F}

```

Problem 2:
```{r a2, ref.label='p2', eval=F}

```

Problem 3:
```{r a3, ref.label='p3', eval=F}

```

```{r a3a, ref.label='p3a', eval=F}

```

```{r a3b, ref.label='p3b', eval=F}

```

Problem 4:
```{r a4, ref.label='p4', eval=F}

```

Problem 5:
```{r a5, ref.label='p5', eval=F}

```


